---
title: '"From Evropa to Gayropa": Exploring the Relationship Between Euroskepticism and the Rejection of Gay Men and Lesbian Women in Serbia'
author: "Vanessa Azañedo"
date: "2026-02-16"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,   
  fig.height = 6,
  out.width = "50%", 
  dpi = 300
)
```

# 1. Data Preparation

## 1.1 Environment Setup

We set up the path, load the necessary libraries, and import the ESS11 round 4 file.

```{r, message = FALSE, warning = FALSE}
#Save the path
path <- "/Users/str8d8/SSDA/SIMM61/IndividualPaper" # Change to your working directory
setwd(path) 

#Load packages
library(tidyverse)
library(readr)
library(ggplot2)
library(broom)
library(skimr)
library(dplyr)
library(psych)
library(forcats)
library(stargazer)
library(webshot2)
library(modelsummary)
library(gt)
library(gtsummary)
library(kableExtra)
library(car)
library(sjPlot)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)

# Import the file
ds <- file.path(path, "ESS11e04_0.csv") # Change it accordingly
ESS11 <- read_csv(ds)
```

## 1.2 Data Preprocessing

The data set is filtered to only consider observations from Serbia.
```{r, message = FALSE, warning = FALSE}
# Filter observations 
ESS11_RS <- ESS11 %>% 
  filter(cntry == "RS")
```

We use a tidyverse pipeline to isolate relevant variables, recode missing values, and delete rows that contain them.
```{r, message = FALSE, warning = FALSE}
# 1. Select and mutate variables using a pipeline
ESS11_RS <- ESS11_RS %>%
  # Select
  select(
    agea, gndr, eduyrs, rlgdgr,      # Control variables
    atcherp, euftf, trstep,                    # Attitudes toward EU
    freehms, hmsfmlsh, hmsacld                 # LGBT Acceptance
  ) %>%
  # 2. Assign NAs 
  mutate(
    # Custom thresholds: agea > 150, eduyrs >= 77
    agea     = if_else(agea > 150 | agea == 999, NA_real_, as.numeric(agea)),
    eduyrs   = if_else(eduyrs >= 77, NA_real_, as.numeric(eduyrs)),
    # Standard ESS 0-10 scales (77, 88, 99 are missing)
    across(c(rlgdgr, atcherp, euftf, trstep), ~if_else(.x >= 77, NA_real_, as.numeric(.x))),
    # Standard ESS 1-5 scales (7, 8, 9 are missing)
    across(c(freehms, hmsfmlsh, hmsacld), ~if_else(.x >= 7, NA_real_, as.numeric(.x))),
    # Gender becomes NA if it is NOT 1 (Male) or 2 (Female)
    gndr     = if_else(!gndr %in% c(1, 2), NA_real_, as.numeric(gndr)),
  ) %>%
  # 3. Drop rows with missing values
  drop_na(agea, gndr, eduyrs, rlgdgr, freehms, hmsfmlsh, hmsacld, atcherp, euftf, trstep) %>%
  # 4. Finalize variable types
mutate(
    gndr = as_factor(gndr),
    across(where(is.logical), as.numeric)
  )
```

## 1.3 Data Overview

We briefly inspect the data.
```{r, message = FALSE, warning = FALSE}
# Inspect the data
# Not using View because it obstructs the knitting process >:(
glimpse(ESS11_RS)
head(ESS11_RS)
tail(ESS11_RS)
```

## 1.4 European Context

To provide context, we process the full European dataset to compare average LG rejection scores across all participating ESS11 countries.
```{r}
# 1. Create clean data set
europe_clean <- ESS11 %>%
  # Handle NAs for the three variables
  mutate(across(c(freehms, hmsacld, hmsfmlsh), ~ifelse(. > 5, NA, .))) %>%
  # Reverse ONLY hmsfmlsh so 1 becomes 5 (High Rejection)
  mutate(hmsfmlsh_rev = 6 - hmsfmlsh) %>%
  # Calculate the index
  mutate(LG_Rejection_Index = rowMeans(across(c(freehms, hmsacld, hmsfmlsh_rev)), na.rm = TRUE)) %>%
  filter(!is.na(LG_Rejection_Index))

# 2. Aggregate by country (The 'country_comparison' object)
country_comparison <- europe_clean %>%
  group_by(cntry) %>%
  summarise(
    rejection_score = weighted.mean(LG_Rejection_Index, w = pspwght, na.rm = TRUE)
  )

# 3. Prepare geography
world <- ne_countries(scale = "medium", returnclass = "sf")
europe_geo <- world %>% filter(continent == "Europe")

# 4 Join using the more reliable 'iso_a2_eh' column since certain countries were not showing up initially
map_final <- europe_geo %>%
  left_join(country_comparison, by = c("iso_a2_eh" = "cntry"))

# 5. Plot with BuPu Palette
ggplot(data = map_final) +
  geom_sf(aes(fill = rejection_score), color = "white", size = 0.1) +
  # BuPu: Light Blue to Deep Purple
  # na.value is set to a very light grey for countries not in ESS11
  scale_fill_distiller(palette = "BuPu", 
                       direction = 1, 
                       name = "Rejection of gays and lesbians",
                       na.value = "#f0f0f0") + 
  # Focus coordinates
  coord_sf(xlim = c(-12, 45), ylim = c(35, 70), expand = FALSE) +
  labs(title = "Rejection of Gays and Lesbians across Europe",
       subtitle = "Purple = Higher rejection | Azure  = Lower rejection",
       caption = "Data: ESS Round 11 | Missing values in light grey") +
  theme_minimal() +
  theme(panel.grid = element_blank()) # Clean up the background
```

# 2. Variable Construction

## 2.1 Euroskepticism Index

We reverse the scale of the EU items to ensure that higher numeric values consistently represent higher levels of rejection.
```{r, message = FALSE, warning = FALSE}
# Reversing EU items so HIGHER scores = HIGHER REJECTION
ESS11_RS <- ESS11_RS %>%
  mutate(
    atcherp_rev = 10 - atcherp,
    euftf_rev   = 10 - euftf,
    trstep_rev  = 10 - trstep
  )
```

We conduct a factor analysis and reliability check to construct a unified latent index for Euroskepticism.
```{r}
# 1. Subset the REVERSED items
EU_items <- ESS11_RS %>% 
  select(atcherp_rev, euftf_rev, trstep_rev)
# 2. Run Factor Analysis
fa_EU <- fa(
  EU_items,
  nfactors = 1, 
  rotate   = "none",      # Rotation is omitted here because we only have 1 factor
  fm       = "ml",        # "ml" = Maximum Likelihood factor extraction method
  scores   = "regression" # Compute factor scores for each respondent using regression scoring 
)
# Display loadings
print(fa_EU$loadings, cutoff = 0.3)
# 3. Append factor scores to the main dataframe
# We use the factor score as our latent variable for "Euroskepticism"
ESS11_RS$Euroskepticism_Index <- as.numeric(fa_EU$scores) 
# 4. Check Reliability
alpha_EU <- psych::alpha(EU_items, check.keys = TRUE)
print(alpha_EU$total$raw_alpha)
```

I check the range to make sure all the NAs have been effectively removed.
```{r}
# Check the range of items
summary(ESS11_RS %>% select(atcherp_rev, euftf_rev, trstep_rev))
```

We visualize the distribution of the Euroskepticism latent factor scores to assess normality and spread.
```{r, message = FALSE, warning = FALSE}
ggplot(ESS11_RS, aes(x = Euroskepticism_Index)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "cornflowerblue", color = "white") +
  geom_density(color = "darkblue", size = 1) +
  labs(
    title = "Distribution of Euroskepticism Index",
    subtitle = "Latent Factor Scores (Mean = 0, SD = 1)",
    x = "Euroskepticism (Factor Score)",
    y = "Density"
  ) +
  theme_minimal()
```

## 2.2 Rejection of Gays and Lesbians Index

We reverse the 'hmsfmlsh' item to align it with the other indicators so that high scores represent increased rejection.
```{r, message = FALSE, warning = FALSE}
# Reverse ONLY hmsfmlsh so 5 = higher rejection for all three items
ESS11_RS <- ESS11_RS %>% 
  mutate(hmsfmlsh_rev = 6 - hmsfmlsh)
```

We use factor analysis to extract a single latent variable representing the rejection of gay men and lesbian women.
```{r}
# 1. Prepare items for REJECTION of Gays and Lesbians
# We want 5 to be the "Anti" end for all.
LG_items <- ESS11_RS %>% 
  select(freehms, hmsacld, hmsfmlsh_rev) 
# 2. Run Factor Analysis
# We extract 1 factor representing the latent "Rejection of Gays and Lesbians" construct
fa_LG <- fa(
  LG_items,
  nfactors = 1, 
  rotate   = "none", 
  fm       = "ml", 
  scores   = "regression"
)
# 3. Append factor scores
# Higher values = Higher REJECTION of Gays and Lesbians
ESS11_RS$LG_Rejection_Index <- as.numeric(fa_LG$scores)
# 4. Check Reliability
alpha_LG <- psych::alpha(LG_items, check.keys = TRUE)
print(alpha_LG$total$raw_alpha)
```

We plot the distribution of the Gays and Lesbians Rejection Index to inspect the frequency of homophobic attitudes in the sample.
```{r}
# Visualization of the Rejection of Gays and Lesbians Index
ggplot(ESS11_RS, aes(x = LG_Rejection_Index)) +
  # Histogram with density scaling
  geom_histogram(aes(y = ..density..), bins = 30,
                 fill = "#F4A3A3",   # light red
                 color = "white",
                 alpha = 0.7) +
  # Density curve overlay
  geom_density(color = "#8B0000", size = 1) +  # dark red
  # Labels
  labs(
    title = "Distribution of Rejection of Gays and Lesbians Index",
    subtitle = "Latent Factor Scores (Mean = 0, SD = 1)",
    x = "Rejection of Gays and Lesbians Index (Factor Score)",
    y = "Density"
  ) +
  theme_minimal()
```

# 3. Regression Models

## 3.1 Initial Models

### Model 1: Focal Bivariate Model

We estimate the baseline bivariate relationship between Euroskepticism and LG rejection through a simple linear regression.
```{r}
# Inspect the relationship between the focal variables
Model1 <- lm(LG_Rejection_Index ~ Euroskepticism_Index, data = ESS11_RS)
summary(Model1)
# Run diagnostics 
plot(Model1)
```

### Model 2: Full Model

We expand the model to include socio-demographic controls to test the focal relationship for potential spuriousness.
```{r}
# We include control variables such as gender, age, years of education and religiousness
Model2 <- lm(LG_Rejection_Index ~ Euroskepticism_Index +
               gndr + 
               agea + 
               eduyrs + 
               rlgdgr, 
             data = ESS11_RS)
summary(Model2)
# Run diagnostics  
plot(Model2)
```

We calculate Variance Inflation Factor (VIF) scores to ensure that the multivariate model does not suffer from multicollinearity.
```{r}
# Check for absence of multicollinearity
vif(Model2)
```

We calculate Cook's Distance to identify and remove influential outliers that might bias the regression results.
```{r}
# Display the cases with the highest numbers of Cook's D
CooksD <- cooks.distance(Model2)
sort(CooksD, decreasing = TRUE) %>% head()

# Delete outliers
ESS11_RS_clean <- ESS11_RS[-c(375, 779, 999), ] 
```

## 3.2 Analytical Sample Descriptives

Having deleted the outlier cases, we create a table with a summary of descriptives for the final analytical sample.
```{r, message = FALSE, warning = FALSE}
# Prepare the analytical sample
# Ensure your path and folders are ready
if (!dir.exists(file.path(path, "outputs"))) {
  dir.create(file.path(path, "outputs"), recursive = TRUE)
}

analytical_sample <- ESS11_RS_clean %>%
  mutate(female = if_else(gndr == 2, 1, 0)) %>%
  select(agea, female, eduyrs) %>%
  as.data.frame()

# Define filenames
html_desc <- file.path(path, "outputs", "demographics.html")
png_desc  <- file.path(path, "outputs", "demographics.png")

# Generate the polished table using the same "stargazer" style
stargazer(
  analytical_sample, 
  type = "html", 
  title = "Demographic Characteristics of the Analytical Sample",
  covariate.labels = c("Age", "Female (Dummy)", "Education (Years)"),
  digits = 2,
  summary.stat = c("n", "mean", "sd", "min", "max"),
  notes = "Source: European Social Survey Round 11 (2023).",
  style = "apsr", # Matches the regression table style
  out = html_desc
)

# Convert to PNG
webshot2::webshot(
  url  = html_desc,
  file = png_desc,
  zoom = 1
)
```

## 3.3 Final Models

After removing the three most influential outliers, we run the models again.

### Final Model 1: Bivariate Model

We re-estimate the bivariate model using the cleaned dataset.
```{r}
# Model 1
Model1_final <- lm(LG_Rejection_Index ~ Euroskepticism_Index, data = ESS11_RS_clean)
summary(Model1_final)
# Run diagnostics 
plot(Model1_final)
```

### Final Model 2: Full Model with Control Variables

Again we re-estimate the full multivariate model on the cleaned data to obtain our final coefficients.
```{r}
# Full model (clean)
Model2_final <- lm(LG_Rejection_Index ~ Euroskepticism_Index +
               gndr + 
               agea + 
               eduyrs + 
               rlgdgr, 
             data = ESS11_RS_clean)
summary(Model2_final)

# Diagnostics 
plot(Model2_final)
```

We perform a final VIF check on the cleaned model to confirm the independence of our predictors.
```{r}
# Check for absence of multicollinearity 
vif(Model2_final)
```
The VIF values for all predictors are very close to 1.0, indicating an absence of multicollinearity and confirming that each variable provides independent information to the model.

## 3.4 Model Comparison

We compare the fit of the bivariate and multivariate models.
```{r}
anova(Model1_final, Model2_final)
```

## 3.5 Visualizations

We generate a scatter plot with a regression overlay to visualize the linear trend between Euroskepticism and rejection of gays and lesbians.
```{r}
# Create a scatter plot with the regression line
ggplot(ESS11_RS_clean, aes(x = Euroskepticism_Index, y = LG_Rejection_Index)) +
  # 1. Add points with 'jitter' to see the density in the spikes
  geom_jitter(alpha = 0.3, color = "cornflowerblue") + 
  # 2. Add the linear regression line
  geom_smooth(method = "lm", color = "darkred", se = TRUE) +
  # 3. Labeling
  labs(title = "Relationship between Euroskepticism and Rejection of Gays and Lesbians",
       subtitle = "Focal Bivariate Model",
       x = "Euroskepticism",
       y = "Rejection of Gays and Lesbians") +
  theme_minimal()
```

We create a marginal effects plot to show the predicted values of LG rejection across the range of Euroskepticism, holding controls constant.
```{r}
# Create the Marginal Effects Plot
plot_model(Model2_final, 
           type = "pred", 
           terms = "Euroskepticism_Index [all]", # Calculate predictions for the full range
           title = "Predicted Rejection of Gays and Lesbians by Euroskepticism Levels",
           axis.title = c("Euroskepticism", "Predicted Rejection of Gays and Lesbians Score")) +
  # Adding a caption to explain the controls
  labs(caption = "Adjusted for Gender, Age, Education, and Religiosity.\nShaded area represents 95% Confidence Interval.") +
  theme_minimal()
```

We examine the influence of our controls with a forest plot.
```{r}
# Define a clean mapping for our variables
# The order here determines the order on the graph (top to bottom)
cm <- c(
  'rlgdgr' = 'Religiosity',
  'eduyrs' = 'Education (Years)',
  'agea'   = 'Age',
  'gndr2'  = 'Female (Dummy)',
  'Euroskepticism_Index' = 'Euroskepticism (Focal IV)'
)

# Generate the plot
modelplot(Model2_final, coef_map = cm) + 
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Predictors of Rejection toward Gays and Lesbians",
    x = "Coefficient Estimate (with 95% Confidence Interval)",
    y = ""
  )
```

Using stargazer, we create a final regression table comparing the bivariate and multivariate models.
```{r, results = "hide"}
# 1. First, make sure your folder is ready
if (!dir.exists(file.path(path, "outputs"))) {
  dir.create(file.path(path, "outputs"), recursive = TRUE)
}

# 2. Then, define your filenames
html_file <- file.path(path, "outputs", "serbia_results.html")
png_file  <- file.path(path, "outputs", "serbia_results.png")

# 3. Now run the stargazer command
stargazer(
  Model1_final, Model2_final,
  type = "html",
  title = "Impact of Euroskepticism on Rejection of Gays and Lesbians in Serbia",
  style = "apsr", # American Political Science Review style
  column.labels = c("Bivariate", "Multivariate"),
  dep.var.labels = "LG Rejection Index",
  covariate.labels = c(
    "Euroskepticism Index",
    "Gender: Female",
    "Age (Years)",
    "Education (Years)",
    "Religiosity",
    "Constant"
  ),
  # Standard significance stars for sociology
  star.cutoffs = c(0.05, 0.01, 0.001), 
  omit.stat = c("ser", "f"), # Omit residual error and F-stat to keep it clean
  notes = "Source: European Social Survey Round 11 (2023).",
  out = html_file
)

# Convert the HTML table to PNG 
webshot2::webshot(
  url  = html_file,
  file = png_file,
  zoom = 2
)
```
